---
title: "Bebop Data - ZK"
author: "Zo Kupelian"
date: "2026-06-24"
output: pdf_document
---

#starter

##packages
```{r}
#install.packages("ggplot2")
#install.packages("dplyr")
#install.packages("usethis")
library(ggplot2)
library(dplyr)
library(maps)
library(usethis)
library(httr2)
library(jsonlite)
library(purrr)
library(lubridate)
```

##load'n'look
```{r}
#establishing characteristics of data 

FET <- read.csv("FET_OK.csv")  #loading in Final Edition Tested dataset from 
                               #Olivia
head(FET)                      #looking at 1st 10 rows
nrow(FET)                      #how many ticks we have data for 

sum(is.na(FET$latitude))       #how many latitude data points we're missing
sum(is.na(FET$longitude))      #how many longitude data points we're missing
```

##transformation
```{r}
#making YES = 1 and NO = 0 for tick_infection and tick_infection_tested

#these two columns are whether or not the tick in question has been tested 
#(tick_infection_tested) and if that test came back positive for lyme disease 
#(tick infection)

FET$tick_infection_tested <- ifelse(FET$tick_infection_tested == "YES", 1, 0)
FET$tick_infection <- ifelse(FET$tick_infection == "YES", 1, 0)

#creating an additional "date" column from collection date information for use
#later when scraping for weather data on a particular date

FET$date <- as.Date(       #useful function, requires day month and year
  paste(
    FET$collection_year,
    FET$collection_month,
    FET$collection_day,
    sep = "-"),
  format = "%Y-%m-%d"
)
```

##exploration
```{r}
#just for ticks that HAVE BEEN TESTED, I want to see across all years which 
#months we see more ticks tested

ticks_by_month_all_graph <- FET %>%
  filter(tick_infection_tested == 1) %>%           #only graphing the ticks
  ggplot(aes(x = factor(collection_month))) +      #that HAVE been tested
  geom_bar() +
  labs(
    x = "Collection Month",
    y = "Ticks Tested",
    title = "Ticks Tested per Month"
  )
ticks_by_month_all_graph

#now just for the ticks that TESTED POSITIVE by month across all years

positives_by_month_all_graph <- FET %>%
  filter(tick_infection == 1) %>%                 #only graphing the ticks
  ggplot(aes(x = factor(collection_month))) +     #that tested positive
  geom_bar() +
  labs(
    x = "Collection Month",
    y = "Ticks Tested Positive",
    title = "Ticks Positive per Month"
  )
positives_by_month_all_graph
```

```{r}
#now generate the same graphs, but by year

ticks_by_year_graph <- FET %>%
  filter(tick_infection_tested == 1) %>%           #only graphing the ticks
  ggplot(aes(x = factor(collection_year))) +      #that HAVE been tested
  geom_bar() +
  labs(
    x = "Collection Year",
    y = "Ticks Tested",
    title = "Ticks Tested per Year"
  )
ticks_by_year_graph

#now just for the ticks that TESTED POSITIVE by month across all years

positives_by_year_graph <- FET %>%
  filter(tick_infection == 1) %>%                 #only graphing the ticks
  ggplot(aes(x = factor(collection_year))) +     #that tested positive
  geom_bar() +
  labs(
    x = "Collection Year",
    y = "Ticks Tested Positive",
    title = "Ticks Positive per Year"
  )
positives_by_year_graph
```

```{r}
#okay, now we'll just do total number of ticks collected by year
total_ticks_by_year_graph <- FET %>%
  ggplot(aes(x = factor(collection_year))) +
  geom_bar() +
  labs(
    x = "Collection Year",
    y = "Ticks Collected",
    title = "Ticks Collected per Year"
  )
total_ticks_by_year_graph
```

It's weird 2022 and 2023 are so low, we'd actually expect to see the opposite. 
Actually, I was expecting to see an increase over time generally. Probably some 
missing data, but this will skew our significance by year.

## ratios

Of the ticks that were tested, what proportion are positive each year? How 
does it change year to year?
```{r}
#annual_summary is just grouping by collection year and adding up all of the 
#ticks collected and the results of the tests to create a "prevalence" column

annual_summary <- FET %>%
  group_by(collection_year) %>%
  summarise(                       #just adding up all positives and all tests
    tested = sum(tick_infection_tested == 1, na.rm = TRUE),
    positive = sum(tick_infection == 1, na.rm = TRUE),
    .groups = "drop"
  ) %>%
  mutate(
    prevalence = positive / tested,
    se = sqrt(prevalence * (1 - prevalence) / tested), #standard error
    lower_ci = pmax(0, prevalence - 1.96 * se), #lower 95% confidence interval
    upper_ci = pmin(1, prevalence + 1.96 * se)  #upper 95% confidence interval
  )

#testing if there's significant difference across years

#measuring the difference in positive vs. negative ticks in cbind() beacuse
#we need the total number of ticks that are tested, which the prevalence 
#variable we just made doesn't give us. Prevalence is just a ratio.

#collection_year has to be a factor, otherwise it'll be treated as a continuous
#numerical variable, and we're not measuring years in like 2022.5 or something

prev_across_years_model <- glm(
  cbind(positive, tested - positive) ~ factor(collection_year),
  data = annual_summary,
  family = binomial
)

drop1(prev_across_years_model, test = "Chisq") #LRT

#these results tell us that prevalence does in fact differ across years
#because this is just prevalence among ticks tested and not dependent on the
#number of ticks collected per year, the weird data discrepancies don't 
#actually invalidate this finding.

#these results will compare prevalence between years, whether it's > or <
summary(prev_across_years_model) #log odds difference between adjacent years

#test for a linear trend over time
prevtrend_across_years_model <- glm(
  cbind(positive, tested - positive) ~ collection_year,
  data = annual_summary,
  family = binomial
)

summary(prevtrend_across_years_model)

#plot positive ratio over time
ggplot(annual_summary, aes(x = collection_year, y = prevalence)) +
  geom_line(group = 1) +
  geom_point(size = 3) +
  geom_errorbar(
    aes(ymin = lower_ci, ymax = upper_ci),
    width = 0.2
  ) +
  scale_y_continuous(
    labels = scales::percent_format(accuracy = 1),
    limits = c(0, 1)
  ) +
  labs(
    x = "Collection Year",
    y = "Prevalence of Tick Infection",
    title = "Annual Tick Infection Prevalence"
  ) +
  theme_bw()
```

##location visual
```{r}
plot(FET_ne$longitude, FET_ne$latitude,
     col = ifelse(FET_ne$tick_infection == 1, "red", "black"),
     pch = 19,
     xlab = "Longitude", ylab = "Latitude")
legend("topright", legend = c("Infected", "Not infected"),
       col = c("red", "black"), pch = 19)
```


#NE case study
```{r}
#going to get rid of any ticks we have that are missing location coords
FET <- FET[!is.na(FET$longitude) & !is.na(FET$latitude), ]

#making a subset of the remaining data that is limited to New England, since
#we have some outliers that are probably just random error and shouldn't factor
#into our analysis

#NOTE: FET_ne IS LIMITING COORDS TO BE IN NEW ENGLAND. WHEN DATA OUTSIDE OF NEW 
#ENGLAND IS ADDED, REMOVE THIS LIMITATION

FET_ne <- subset(
  FET,
  longitude >= -73.7 & longitude <= -66.9 &  #these coords include all of
  latitude  >= 41.0  & latitude  <= 47.5     #the NE states, can be expanded
)

#making a map of coord distribution on just the ne_states

states_map <- map_data("state") #built-in US state border coordinates for 
                                #ggplot
ne_states <- subset(            #picking just the NE states
  states_map,
  region %in% c(
    "massachusetts", "maine", "vermont", "new hampshire", "rhode island", 
    "connecticut"
  )
)

all_time_NE <- ggplot() +
  geom_polygon(                             #geom_polygon, weird state shapes
    data = ne_states,                       #need state shape data from maps
    aes(x = long, y = lat, group = group),  #ne_states has these columns
    fill = "grey25",                        #color of states
    color = "white"                         #color of state borders
  ) +
  geom_point(                               #now plot collection locations
    data = FET_ne,
    aes(x = longitude, y = latitude, color = factor(tick_infection)),
    alpha = 0.5,
    size = 1.5
  ) + 
  coord_fixed(1.3) +   #corrects aspect ratio, looks squished otherwise
  labs(
    color = "Positive Tests",
    title = "Dist. in New England (All Time)"
  )
all_time_NE
```

##NE by year
###note: must run graph_NE_year.R script also included in the download
```{r}
source("graph_NE_year.R") #wrote a function to generate a graph for a 
                          #particular year that would be similar to all_time_NE

#graph_NE_year(FET, 2017)
#graph_NE_year(FET, 2018)
graph_NE_year(FET, 2019)
#graph_NE_year(FET, 2020)
#graph_NE_year(FET, 2021)
#graph_NE_year(FET, 2022)
#graph_NE_year(FET, 2023)
```

##significance
```{r}
#testing significance of year on if a tick is infected or not

ne_year_model <- glm(tick_infection ~ collection_year,
                     data = FET_ne,
                     family = binomial)
ne_year_model
```
We are seeing a negative statistically significant correlation between year and 
positive ticks, but this is probably from overall tick counts. Probably look
into proportion of ticks tested positive would be a better reflection of 
the effect year has on lyme disease prevalence.

```{r}
#same as above, just making the prevalence column for FET_ne specifically
annual_summary_ne <- FET_ne %>%
  group_by(collection_year) %>%
  summarise(                       #just adding up all positives and all tests
    tested = sum(tick_infection_tested == 1, na.rm = TRUE),
    positive = sum(tick_infection == 1, na.rm = TRUE),
    .groups = "drop"
  ) %>%
  mutate(
    prevalence = positive / tested,
    se = sqrt(prevalence * (1 - prevalence) / tested), #standard error
    lower_ci = pmax(0, prevalence - 1.96 * se), #lower 95% confidence interval
    upper_ci = pmin(1, prevalence + 1.96 * se)  #upper 95% confidence interval
  )

#again, recreating the prevalence plot across years but just for New England
ggplot(annual_summary_ne, aes(x = collection_year, y = prevalence)) +
  geom_line(group = 1) +
  geom_point(size = 3) +
  geom_errorbar(
    aes(ymin = lower_ci, ymax = upper_ci),
    width = 0.2
  ) +
  scale_y_continuous(
    labels = scales::percent_format(accuracy = 1),
    limits = c(0, 1)
  ) +
  labs(
    x = "Collection Year",
    y = "Prevalence of Tick Infection",
    title = "Annual Tick Infection Prevalence in New England"
  ) +
  theme_bw()

#model to test that significance across the years, but probably slope is close 
#to 0 just based on the visualization we did previously. Might be statistically
#significant, though.

ne_prevalence_year_model <- glm(prevalence ~ collection_year,
                     data = annual_summary_ne,
                     family = binomial)
ne_prevalence_year_model

#is not statistically significant :(
```

The proportion here of ticks testing positive per ticks tested total is not 
statistically significant across years :(, but has a vague upward trend. With
greater data presence in 2022 and 2023, conclusion and significance may change.

Maybe if I break it down month by month to make monthly_summary_ne I can test
the significance of prevalence by monthly weather

##weather scraping
```{r}
#have to remove any ticks missing latitude, longitude, or date

FET_clean <- FET_ne %>%
  filter(
    !is.na(latitude),
    !is.na(longitude),
    !is.na(date)
  )
nrow(FET_clean) #cleans out 149 ticks of missing data

#makes a new data frame with the info for weather scraping
weather_keys <- FET_clean %>%
  distinct(latitude, longitude, date) #replace distinct(), it's removing
                                      #duplicate points that we want
#load in my custom function scrape_weather
source("scrape_weather.R")

#makes that dataframe into a list
weather_list <- vector("list", nrow(weather_keys))

#----------------------------------------------------------------------
for (i in seq_len(nrow(weather_keys))) {

  if (i %% 100 == 0) {
    message("Completed ", i, " of ", nrow(weather_keys), " weather requests")
  }
#----------------------------------------------------------------------
#everything in the above chunk is a sanity check
#this stuff needs optimizing, basically here to tell me how long below takes
  
  weather_list[[i]] <- scrape_weather(
    latitude = weather_keys$latitude[i],
    longitude = weather_keys$longitude[i],
    date = weather_keys$date[i]
  )

  #Sys.sleep(0.2)   #to avoid hammering the API, this is the bottleneck
}
weather_df <- dplyr::bind_rows(weather_list)
head(weather_df)

#this is returning everything as a list, so I need to unlist it
weather_df <- weather_df %>%
  mutate(across(
    where(is.list),
    ~ unlist(.x)
  ))
head(weather_df)
```

##weather data analysis
```{r}
#combining the weather data with the tick data by matching coordinates and
#date, can use for significance testing now.

FET_ne_weather <- FET_ne %>%
  left_join(weather_df, 
            b = c("longitude", "latitude", "date"))
head(FET_ne_weather)

#linear model for tick infection prevalence and temp/precip/humidity?
```

##monthly weather
```{r}
#this is something I might need to test monthly weather and prevalence 
#eventually, but for the actual monthly heat mapping, I'm just going to pull
#average weather data by month and upload that in layers to arcgis

monthly_summary_ne <- FET_ne_weather %>%
  mutate(
    month = floor_date(date, "month")
  ) %>%
  group_by(location_county, month) %>%
  summarise(
    # Monthly average weather
    temp_mean = mean(temperature_mean, na.rm = TRUE),
    temp_max = mean(temperature_max, na.rm = TRUE),
    temp_min = mean(temperature_min, na.rm = TRUE),
    humidity_mean = mean(humidity_mean, na.rm = TRUE),
    precipitation_sum = mean(precipitation_sum, na.rm = TRUE),
    rain_sum = mean(rain_sum, na.rm = TRUE),
    snowfall_sum = mean(snowfall_sum, na.rm = TRUE),
    precipitation_hours = mean(precipitation_hours, na.rm = TRUE),

    # Tick counts
    ticks_tested = sum(!is.na(tick_infection_tested)),
    ticks_positive = sum(tick_infection == 1, na.rm = TRUE),

    .groups = "drop"
  )
monthly_summary_ne
```

##just weather
```{r}
library(tigris)
library(sf)
library(dplyr)

# New England state abbreviations
ne_states <- c("CT", "ME", "MA", "NH", "RI", "VT")

county_locations <- purrr::map_dfr(ne_states, function(st){

  counties(state = st, cb = TRUE, year = 2020) |>
    st_transform(4326) |>             # WGS84 lat/lon
    mutate(centroid = st_centroid(geometry)) |>
    mutate(
      longitude = st_coordinates(centroid)[,1],
      latitude  = st_coordinates(centroid)[,2]
    ) |>
    st_drop_geometry() |>
    transmute(
      location_county = NAME,
      state = STUSPS,
      latitude,
      longitude
    )

})
county_locations


```

