---
title: "MA_cases"
author: "Zo Kupelian"
date: "2026-08-04"
output: pdf_document
---

#libraries
```{r}
#install.packages("readxl")
#install.packages("magick")
library(tidyverse)
library(readxl)
library(dplyr)
library(ggplot2)
library(tigris)
library(sf)
library(stringr)
library(magick)
```

#load data
```{r}
MA_cases <- read_excel("Lyme_by_County_2000_2025.xlsx")
MA_cases
```

I want to make a map of total cases in each county in 2000, to begin with.

#little data cleaning
```{r}
# <5 was used for de-identification, but because that's a character, I'm replacing it with 2.5

MA_cases <- lapply(MA_cases, function(x) gsub("<5", "2.5", x))
MA_cases <- as.data.frame(MA_cases)

MA_cases <- MA_cases %>%
  mutate(
    Year = as.numeric(Year),
    January = as.numeric(January), 
    February = as.numeric(February),
    March = as.numeric(March), 
    April = as.numeric(April), 
    May = as.numeric(May),
    June = as.numeric(June), 
    July = as.numeric(July), 
    August = as.numeric(August),
    September = as.numeric(September), 
    October = as.numeric(October), 
    November = as.numeric(November), 
    December = as.numeric(December)
  )

#making an extra column with the total cases for each county summed over the year
MA_cases$year_total <- rowSums(MA_cases[, c("January", "February", "March", "April", "May", "June", "July", "August", "September", "October", "November", "December")], na.rm = TRUE)

MA_cases %>%
  rename("County_Name" = "County")

county_map <- counties(state = "MA", cb = TRUE, year = 2000) %>%
  rename("County_Name" = "NAME")
```

#heatmapping function

See the comments for standardized vs. not standardized

```{r}
#make a directory to put the heatmaps in
dir_out <- file.path(".", "heatmaps_folder")
dir.create(dir_out, recursive = TRUE)

#note with this function! All of the cases need to be in numeric format

year_heatmap <- function(year_choice) {

#year by year, so filter the data by the year the user wants to see
  MA_cases_year <- MA_cases %>%
    filter(Year == year_choice) %>%
    rename("County_Name" = "County") #rename for joining

#MA counties haven't changed from 2000, so we can use the data from 2000
county_map <- counties(state = "MA", cb = TRUE, year = 2000)
county_map <- county_map %>%
  rename("County_Name" = "NAME") #rename for joining

#Barnstable County --> Barnstable in the MA_cases_year data frame
MA_cases_year$County_Name <- str_replace_all(MA_cases_year$County_Name, " County", "")

#join
merged_data <- left_join(county_map, MA_cases_year, by = "County_Name")

#standardizing across the entire range of case counts
case_limits <- c(0, 2000) #max cases per year in one county is ~1700

#-------------------------------------------------------------------------------
#remove case limits from above and below if we want to see counties compare to
#each other in a certain year

#keep case limits if we're looking across the whole 25-year-long period of time
#-------------------------------------------------------------------------------

#time to plot
heatmap <- ggplot(merged_data) +
  geom_sf(aes(fill = year_total), color = "white", linewidth = 0.2) +
  scale_fill_viridis_c(limits = case_limits, option = "plasma", name = "Number of Cases") +
  theme_void() +
  labs(title = paste("MA Cases per County Heatmap in ", year_choice))
print(heatmap)

fp <- file.path(dir_out, paste0("MA_case_heatmap_", year_choice, ".png"))

#time to save
ggsave(plot = heatmap, filename = fp, device = "png")

}

#uncomment below to get all the heatmaps for all the years

#lapply(2000:2025, year_heatmap)
```

#make a gif
```{r}
# listing the file names, and reading them in

heatmaps <- list.files(dir_out, full.names = TRUE)
heatmaps_list <- lapply(heatmaps, image_read)

#if you can't beat 'em, join 'em
heatmaps_joined <- image_join(heatmaps_list)

#animate!
heatmaps_animated <- image_animate(heatmaps_joined, fps = 2)
heatmaps_animated

#save the gif
image_write(image = heatmaps_animated, path = "MA_cases_heatmap.gif")
```

So, the fact of a 5-year-long inconsistency in a 25-year-long collection window
makes analysis kind of challenging, but that doesn't mean we're not going to try.

I think first things first we'll need to graph each year on an x axis and number of cases on a y axis, and make a vertical stack of all the counties' case counts in that year, differentiated by color. This'll show the trend, which I expect will have a close-to-exponential increase.

Can run models.

#analysis
```{r}
#overall plot
cases_plot <- ggplot() +
  geom_point(data = MA_cases, aes(x = Year, y = year_total, color = County)) +
  geom_smooth(data = MA_cases, aes(x = Year, y = year_total), method = "glm", color = "black", se = TRUE)
cases_plot

#again, this is super screwed by the 2016-2021 data...

#pivot longer for year-by-year visualization
MA_cases_longer <- pivot_longer(
    data = MA_cases,
    cols = c("January", "February", "March", "April", "May", "June", "July", "August", "September", "October", "November", "December"),
    names_to = "Month",
    values_to = "Cases"
    ) %>%
  select("County", "Month", "Cases", "year_total", "Year")
MA_cases_longer

#function to make a graph each year
cases_by_county_graph <- function(year){
  
  MA_cases_year <- MA_cases_longer %>%
    filter(Year == year)

cases_plot_year <- ggplot() +
  geom_point(data = MA_cases_year, aes(x = Month, y = Cases, color = County)) +
  ylim(0, 450) +
  labs(title = paste0("Monthly Cases per County in ", year))
print(cases_plot_year)
}

#uncomment the line below to get the trend yearly
#lapply(2000:2025, cases_by_county_graph)
```

Missing data for 2016-2021 needs to be imputed, which is not ideal.

Looks like polynomial regression to me - not quite linear. Random forest regression?

#--------------------------------------------------------------------------------
#Until imputation is done correctly, everythign will be done on yearly basis,NOT a monthly one. Monthly analysis will come after 2016-2021 have data.
#--------------------------------------------------------------------------------

#NOT IMPUTED
```{r}
model1 <- lm(year_total ~ Year + County, data = MA_cases)
summary(model1)
```

Year is obviously significant, even with the diminished counts for 2016-2021

Dukes, Franklin, Hampshire, Middlesex, Nantucket, Plymouth, Suffolk, and Worcester counties
all see significance in the linear model, though some of them are positive and some are negative, probably because 
of the five-year-issue (FYI).

#Imputation
```{r}
#cleaning out the year_total dates for 2016-2021 so that they can be imputed
MA_cases_pre <- MA_cases

for (i in 2016:2021) {
  MA_cases_pre$year_total[MA_cases_pre$Year == i] <- NA
}
MA_cases_pre
```

```{r}
df <- data.frame(
  id = 1:4,
  category = c("A", "B", "C", "A"),
  score = c(10, 20, 30, 40)
)

df$score[df$category == "A"] <- NA

print(df)
class(df)
```


