---
title: "ml_intro"
author: "Zo Kupelian"
date: "2026-08-31"
output: pdf_document
---

#Option 1

```{r}
#install.packages("tidymodels")
#install.packages("ranger")
#install.packages("leaflet")
library(leaflet)
library(tidymodels)
library(ranger)
library(ggplot2)
library(raster)
library(sp)
```

split for test and train
```{r}
#for reproducability

set.seed(94)

#initial_split is a useful way to take a chunk for training
data_split <- initial_split(iris, prop = 0.8, strata = Species)

train_data <- training(data_split)
test_data <- testing(data_split)
```

Going to try to predict species based on all of the other variables in iris
```{r}
iris_recipe <- recipe(Species ~ ., data = train_data) %>%
  step_normalize(all_numeric_predictors()) #centering and scaling numeric data
```

```{r}
#defining a random forest classification model

rf_spec <- rand_forest(trees = 100) %>%
  set_engine("ranger") %>%
  set_mode("classification")
```

Creating machine learning workflow
```{r}
iris_workflow <- workflow () %>%
  add_recipe(iris_recipe) %>%
  add_model(rf_spec)
```

train and evaluate model
```{r}
final_fit <- last_fit(iris_workflow, data_split)

collect_metrics(final_fit)
```

```{r}
test_predictions <- collect_predictions(final_fit)
test_predictions %>%
  conf_mat(truth = Species, estimate = .pred_class)
```

#Option 2

following the guide here:

https://shittuolumide.medium.com/machine-learning-basics-building-your-first-predictive-model-in-r-010ffe88d6e4

This uses caret and tidyverse, which implies that it might be updated to use tidymodels instead

Going to use mtcars for the first attempt, then try with iris later

```{r}
#install.packages("tidyverse")
#install.packages("caret")
library(tidyverse)
library(caret)
```

```{r}
data <- mtcars
data

#look for missingness
sum(is.na(data))

#summary stats for all the columns
summary(data)

#split the data into 80% training and 20% testing
train_index <- createDataPartition(data$mpg, p = 0.8, list = FALSE)
train_data <- data[train_index, ]
test_data <- data[-train_index, ]

#train the model, using the train() function from caret
#this is a linear regression model, hence the use of "lm"

model <- train(
  mpg ~ wt + hp, #forumula, mpg as a function of wt and hp
  data = train_data,
  method = "lm",
  trControl = trainControl(method = "cv", number = 5) #five-fold cross validation
)

print(model)
```

Evaluating the model
```{r}
#make predictions
predictions <- predict(model, newdata = test_data)

results <- data.frame(
  Actual = test_data$mpg,
  Predicted = predictions
)

#root mean square error (RMSE)
rmse <- sqrt(mean((results$Actual - results$Predicted)^2))
print(paste("RMSE:", rmse))
```

Not bad! Let's graph it for fun
```{r}
plot(test_data$mpg, predictions, main = "Actual vs. Predicted",
     xlab = "Actual mpg", ylab = "Predicted mpg")
abline(0, 1, col = "red")
```

##recreate
Okay, let's see if I understand this enough to recreate if for iris

```{r}
sum(is.na(iris))
summary(iris)

train_iris_index <- createDataPartition(iris$Petal.Length, p = 0.8, list = FALSE)
train_iris_data <- iris[train_iris_index, ]
test_iris_data <- iris[-train_iris_index, ]
```

```{r}
#training a model to predict petal length given sepal width and length

iris_model <- train(
  Petal.Length ~ Sepal.Width + Sepal.Length,
  data = train_iris_data,
  method = "lm",
  trControl = trainControl(method = "cv", number = 5) #this specifically I need to clarify for myself
)

print(iris_model)
```

```{r}
iris_predictions <- predict(iris_model, newdata = test_iris_data)

iris_results <- data.frame(
  Iris_Actual = test_iris_data$Petal.Length,
  Iris_Predicted = iris_predictions
)

iris_rmse <- sqrt(mean((iris_results$Iris_Actual - iris_results$Iris_Predicted)^2))

print(paste("RMSE:", iris_rmse))
```

Visualize!
```{r}
plot(test_iris_data$Petal.Length, iris_predictions, main = "Actual vs. Predicted Petal Length",
     xlab = "Actual petal length",
     ylab = "Predicted petal length")
abline(0, 1, col = "red")
```

##real data

Okay, looks like I can. That's pretty cool. Maybe I'll give it a shot with the app data.

```{r}
#install.packages("xlsx")
library(xlsx)
library(tidyverse)
library(maps)
library(ggplot2)
library(dplyr)
```

```{r}
app_data <- read.xlsx("app_data.xlsx", sheetIndex = 1)
app_data <- app_data %>%
  select(created_at_utc, latitude, longitude, result, readable, confidence, test_line, invalid_reason, user_action, reasoning, parsed_result, parsed_confidence, parsed_control_line, parsed_test_line, parsed_invalid_reason)

app_data
```

##exploration
```{r}
world_map <- map_data("world")

ticks <- data.frame(lon = as.numeric(c(app_data$longitude)), lat = as.numeric(c(app_data$latitude)))

world_map_graphed <- ggplot() +
  geom_polygon(data = world_map, aes(x = long, y = lat, group = group), fill = "grey", color = "white", linewidth = 0.2) +
  geom_point(data = ticks, aes(x = lon, y = lat), color = "red", size = 1)
world_map_graphed
```

Wow, that's kind of interesting, I wasn't expecting to have some app users in India and Singapore

They'll make for some interesting outliers.

I'll test just using latitude and longitude for now, but I'll set up weather pulls from an API later. Also figure out a way to include date, with later/sooner as the options

##data cleaning
```{r}
#invalids are supposed to be such a small percentage of the dataset in the future that removing them shouldn't impact the model
#though, with this starting dataset they are a kind of huge portion

sum(app_data$result == "invalid", na.rm = TRUE) #76 out of 182 samples are invalid. yikes.

#filtering out invalid results
app_data_clean <- app_data %>%
  filter(!grepl("invalid", result, ignore.case = TRUE)) %>%
  mutate(across(where(is.character), ~ str_replace_all(., c("positive" = "1", "negative" = "0"))))

app_data_clean$result <- as.numeric(app_data_clean$result)
app_data_clean$latitude <- as.numeric(app_data_clean$latitude)
app_data_clean$longitude <- as.numeric(app_data_clean$longitude)
app_data_clean
```


```{r}
set.seed(911)

train_tick_index <- createDataPartition(app_data_clean$result, p = 0.8, list = FALSE)
train_tick_data <- app_data_clean[train_tick_index, ]
test_tick_data <- app_data_clean[-train_tick_index, ]
```

```{r}
tick_model <- train(
  result ~ latitude + longitude,
  data = train_tick_data,
  method = "lm",
  trControl = trainControl(method = "cv", number = 5) #figure out what this is!!
)

print(tick_model)
```

```{r}
tick_predictions <- predict(tick_model, newdata = test_tick_data)

tick_results <- data.frame(
  Tick_Actual = test_tick_data$result,
  Tick_Predicted = tick_predictions
)

tick_rmse <- sqrt(mean((tick_results$Tick_Actual - tick_results$Tick_Predicted)^2))
print(paste("RMSE:", tick_rmse))
```

```{r}
plot(test_tick_data$result, tick_predictions, main = "Actual vs. Predicted Testing Result",
     xlab = "Actual test result",
     ylab = "Predicted test result"
)
abline(0, 1, col = "blue")
```

So! The result of this model isn't great! But I'd have been unbelievably 
surprised if it was considering it's using linear regression for a binary,
spatial set of data.

Next steps are going to be swapping "lm" out for something else, and getting 
weather data to provide that model.

##take 2

```{r}
#set result to a factor for glm with caret, this is a classification model
app_data_clean$result <- as.factor(app_data_clean$result)

set.seed(911)

#80% training, 20% testing. Mostly bc the number of data points we have is small

#stratified by result, maintaining the same class proportion in both sets so we
#don't end up with a testing set full of negatives

train_tick_index <- createDataPartition(app_data_clean$result, p = 0.8, list = FALSE)

#all this is is an index of row numbers for the training set. Just numbers.

train_tick_data <- app_data_clean[train_tick_index, ]
test_tick_data <- app_data_clean[-train_tick_index, ]
```

```{r}
#uses caret::train() function to train the model, may want to look at alternatives

#result (positive or negative) predicted by latitude and longitude, using only
#geographic coordinates to classify the result

tick_model <- train(
  result ~ latitude + longitude,
  data = train_tick_data,
  method = "glm", #generalized linear regression model
  family = binomial(link = "logit"), #specifying logistic regression, binary outcome and outputting probs 0 to 1
  trControl = trainControl(method = "cv", number = 5) #5-fold cross-validation during training
)

#What does that trControl mean? It divides the training data into 5 subsets, trains 5 times, each time using 4 folds and 1 to validate
#you can increase this, but the more subsets you divide the training data into, the smaller they get. Diminishing returns and all that.

print(tick_model)
#coefficients, performance metrics, sample sizes, etc.
```

An accuracy of 62.8% percent with the 5-fold validation isn't amazing, but it's 
better than flipping a coin!

Below is predictions and confusion matrix
```{r}
#these are probability predictions on the test set, taking the test data and 
#using it to predict "result = 1" or "result = 0"

#eventually we'll compare and see if it matches up to the *actual* "result = 1" or "result = 0"

#P(result = 1 | latitude, longitude)

tick_predictions_prob <- predict(tick_model, newdata = test_tick_data, type = "prob") #type = "prob" is probabilities for each class
tick_predictions <- predict(tick_model, newdata = test_tick_data, type = "raw") #0 or 1 is returned as the predicted class

#making that dataframe of the actual test data and the predicted test data
tick_results <- data.frame(
  Tick_Actual = test_tick_data$result,
  Tick_Predicted = tick_predictions
)

#confusion matrix to show true positives, true negatives, false positives, false 
#negatives, accuracy, sensitivity, specificity, etc.
confusionMatrix(tick_predictions, test_tick_data$result)
```

70% accuracy, so not terrible. Good sensitivity, specificity could be better.

```{r}
#this is visualization

# the `1` extracts the probability column for the positive class

tick_results <- data.frame(
  Tick_Actual = test_tick_data$result,
  Tick_Predicted = tick_predictions,
  Probability = tick_predictions_prob$`1`  # Predicted probability
)

# Confusion Matrix Visualization

#x axis shows the actual results, the fill colors is the predicted results

ggplot(tick_results, aes(x = Tick_Actual, fill = Tick_Predicted)) +
  geom_bar(position = "dodge") +
  labs(title = "Actual vs. Predicted Classifications",
       x = "Actual Result", y = "Count") +
  theme_minimal()
```

The model does a great job predicting when ticks are going to be negative based
on location data. It does a less great job when ticks are going to be positive
based on location data.

Okay, let's attempt another visualization where I have on map with the test
points in a color, then color for the predicted points.

```{r}
app_data_clean

```


#Tick movement

```{r}
#parsing datetime from the created_at_utc column

#recognizing the current format of the date column needs to be adjusted to 
#as.POSIXct format
app_data_clean$created_at_utc <- as.POSIXct(app_data_clean$created_at_utc,
                                            format = "%Y-%m-%dT%H:%M:%OS",
                                            tz = "UTC")

#new cleaning of data
app_data_spatio <- app_data_clean %>%
  select(created_at_utc, latitude, longitude, result, readable, parsed_result, parsed_confidence) %>%
  arrange(created_at_utc) %>%
  mutate(
    # Temporal features
    day_of_year = yday(created_at_utc),
    month = month(created_at_utc),
    week_of_year = week(created_at_utc),
    hour = hour(created_at_utc),
    date_numeric = as.numeric(created_at_utc),  # Days since epoch
    
    # Coordinate changes over time (velocity/movement)
    lat_change = latitude - lag(latitude),
    lon_change = longitude - lag(longitude),
    coord_distance = sqrt(lat_change^2 + lon_change^2),  # Euclidean distance moved
    lat_change_ma = zoo::rollmean(lat_change, k = 5, fill = NA),  # 5-point moving avg
    lon_change_ma = zoo::rollmean(lon_change, k = 5, fill = NA)
  ) %>%
  
  #small problem here, makes the first and last two rows NAs, which the model can't handle, so we're going to back/forward fill
  
  tidyr::fill(lat_change, lon_change, lat_change_ma, lon_change_ma, .direction = "down") %>%
  tidyr::fill(lat_change, lon_change, lat_change_ma, lon_change_ma, .direction = "up") %>%
  filter(!is.na(lat_change) & !is.na(lon_change)) #removing NAs from lagging
app_data_spatio

set.seed(925)

# Remove any remaining NAs from all columns used in the model
train_coord_index <- createDataPartition(app_data_spatio$day_of_year, p = 0.8, list = FALSE)
train_coord_data <- app_data_spatio[train_coord_index, ] %>%
  na.omit()  # Remove any row with ANY NA value

test_coord_data <- app_data_spatio[-train_coord_index, ] %>%
  na.omit()

# Check how many rows remain
cat("Training rows:", nrow(train_coord_data), "\n")
cat("Testing rows:", nrow(test_coord_data), "\n")

# Lat model
lat_model <- train(
  latitude ~ day_of_year + month + week_of_year + hour + lat_change_ma + lon_change_ma + coord_distance,
  data = train_coord_data,
  method = "lm",
  trControl = trainControl(method = "cv", number = 5)
)

# Lon model
lon_model <- train(
  longitude ~ day_of_year + month + week_of_year + hour + lat_change_ma + lon_change_ma + coord_distance,
  data = train_coord_data,
  method = "lm",
  trControl = trainControl(method = "cv", number = 5) 
)

print(lat_model)
print(lon_model)
```

Make predictions on the test set
```{r}
lat_predictions <- predict(lat_model, newdata = test_coord_data)
lon_predictions <- predict(lon_model, newdata = test_coord_data)

#compare actual vs. predicted coordinates
coord_results <- data.frame(
  actual_latitude = test_coord_data$latitude,
  predicted_latitude = lat_predictions,
  actual_longitude = test_coord_data$longitude,
  predicted_longitude = lon_predictions,
  lat_error = test_coord_data$latitude - lat_predictions,
  lon_error = test_coord_data$longitude - lon_predictions
)

# root mean squared error!
rmse_lat <- sqrt(mean(coord_results$lat_error^2))
rmse_lon <- sqrt(mean(coord_results$lon_error^2))

cat("Latitude RMSE: ", rmse_lat, "\n")
cat("Longitude RMSE: ", rmse_lon, "\n")

#latitude
ggplot(coord_results, aes(x = actual_latitude, y = predicted_latitude)) +
  geom_point(alpha = 0.5) +
  geom_abline(intercept = 0, slope = 1, color = "blue", linetype = "dashed") +
  labs(title = "Predicted vs. Actual Latitude",
       x = "Actual Latitude",
       y = "Predicted Latitude") +
  theme_minimal()

#longitude
ggplot(coord_results, aes(x = actual_longitude, y = predicted_longitude)) +
  geom_point(alpha = 0.5) +
  geom_abline(intercept = 0, slope = 1, color = "red", linetype = "dashed") +
  labs(title = "Predicted vs. Actual Longitude",
       x = "Actual Longitude", y = "Predicted Longitude") +
  theme_minimal()
```

#Predicting Tick Location

##annotate and understand

```{r}
# Create a dataframe with predicted coordinates and additional info
predicted_ticks <- data.frame(
  predicted_latitude = coord_results$predicted_latitude,
  predicted_longitude = coord_results$predicted_longitude,
  actual_latitude = coord_results$actual_latitude,
  actual_longitude = coord_results$actual_longitude,
  lat_error = coord_results$lat_error,
  lon_error = coord_results$lon_error,
  prediction_error = sqrt(coord_results$lat_error^2 + coord_results$lon_error^2),
  datetime = test_coord_data$created_at_utc,
  day_of_year = test_coord_data$day_of_year,
  month = test_coord_data$month,
  result = test_coord_data$result,
  parsed_result = test_coord_data$parsed_result,
  parsed_confidence = test_coord_data$parsed_confidence
)

head(predicted_ticks)

# Basic world map with predicted tick locations
ggplot(predicted_ticks, aes(x = predicted_longitude, y = predicted_latitude)) +
  borders("world", colour = "gray85", fill = "gray80") +
  geom_point(aes(color = prediction_error), alpha = 0.6, size = 2) +
  scale_color_gradient(low = "green", high = "red", name = "Prediction Error") +
  labs(title = "Predicted Tick Coordinates Worldwide",
       x = "Longitude",
       y = "Latitude",
       subtitle = "Green = low error, Red = high error") +
  theme_minimal() +
  coord_fixed()

# Alternative: Color by month (seasonal pattern)
ggplot(predicted_ticks, aes(x = predicted_longitude, y = predicted_latitude)) +
  borders("world", colour = "gray85", fill = "gray80") +
  geom_point(aes(color = factor(month)), alpha = 0.6, size = 2) +
  labs(title = "Predicted Tick Coordinates by Month",
       x = "Longitude",
       y = "Latitude",
       color = "Month") +
  theme_minimal() +
  coord_fixed()

# Alternative: Color by result (positive vs negative)
ggplot(predicted_ticks, aes(x = predicted_longitude, y = predicted_latitude)) +
  borders("world", colour = "gray85", fill = "gray80") +
  geom_point(aes(color = result), alpha = 0.6, size = 2) +
  labs(title = "Predicted Tick Coordinates by Result",
       x = "Longitude",
       y = "Latitude",
       color = "Result") +
  theme_minimal() +
  coord_fixed()
```
#Map Style

```{r}
# Filter to a reasonable sample if you have too many points
predicted_ticks_sample <- predicted_ticks %>%
  sample_n(min(1000, nrow(predicted_ticks)))

# Create color palette based on prediction error
pal <- colorNumeric(palette = "RdYlGn", domain = predicted_ticks_sample$prediction_error, reverse = TRUE)

# Interactive map
leaflet(predicted_ticks_sample) %>%
  addTiles() %>%  # Add OpenStreetMap background
  addCircleMarkers(
    lng = ~predicted_longitude,
    lat = ~predicted_latitude,
    radius = 5,
    color = ~pal(prediction_error),
    stroke = FALSE,
    fillOpacity = 0.7,
    popup = paste("Predicted Lat:", round(predicted_ticks_sample$predicted_latitude, 4), "<br>",
                  "Predicted Lon:", round(predicted_ticks_sample$predicted_longitude, 4), "<br>",
                  "Error:", round(predicted_ticks_sample$prediction_error, 4))
  ) %>%
  addLegend(pal = pal, values = ~prediction_error, title = "Prediction Error")
```

Let's calculate and show a tick risk score!

```{r}
# Grid resolution: 0.5 degrees
grid_resolution <- 0.5

# Create a comprehensive dataframe with both predicted and actual coordinates
risk_data <- data.frame(
  longitude = c(predicted_ticks$predicted_longitude, predicted_ticks$actual_longitude),
  latitude = c(predicted_ticks$predicted_latitude, predicted_ticks$actual_latitude),
  confidence = c(predicted_ticks$parsed_confidence, predicted_ticks$parsed_confidence),
  source = c(rep("predicted", nrow(predicted_ticks)), rep("actual", nrow(predicted_ticks)))
)

# Remove any NA values and convert confidence to numeric
risk_data <- risk_data %>% 
  mutate(confidence = as.numeric(confidence)) %>%
  filter(!is.na(longitude) & !is.na(latitude) & !is.na(confidence))

# Assign each point to a grid cell using floor division (more robust)
risk_data <- risk_data %>%
  mutate(
    grid_lon = floor(longitude / grid_resolution) * grid_resolution,
    grid_lat = floor(latitude / grid_resolution) * grid_resolution
  )

# Calculate risk score: sum of parsed_confidence per grid cell
risk_scores <- risk_data %>%
  group_by(grid_lon, grid_lat) %>%
  summarise(
    risk_score = sum(confidence, na.rm = TRUE),
    point_count = n(),
    avg_confidence = mean(confidence, na.rm = TRUE),
    .groups = 'drop'
  )

# Create global grid
lat_min <- -90
lat_max <- 90
lon_min <- -180
lon_max <- 180

lat_breaks <- seq(lat_min, lat_max, by = grid_resolution)
lon_breaks <- seq(lon_min, lon_max, by = grid_resolution)

grid_cells <- expand.grid(
  grid_lon = lon_breaks,
  grid_lat = lat_breaks
)

# Join with full grid to fill in zeros
risk_grid <- grid_cells %>%
  left_join(risk_scores, by = c("grid_lon" = "grid_lon", "grid_lat" = "grid_lat")) %>%
  mutate(
    risk_score = replace_na(risk_score, 0),
    point_count = replace_na(point_count, 0),
    avg_confidence = replace_na(avg_confidence, 0)
  )

# Create rectangles for each grid cell
risk_grid <- risk_grid %>%
  mutate(
    lon_min = grid_lon - grid_resolution/2,
    lon_max = grid_lon + grid_resolution/2,
    lat_min = grid_lat - grid_resolution/2,
    lat_max = grid_lat + grid_resolution/2
  )

# Normalize risk score for color mapping (0-1)
risk_grid$risk_normalized <- scales::rescale(risk_grid$risk_score, to = c(0, 1))

# Create color palette using viridis
viridis_pal <- leaflet::colorNumeric(
  palette = "viridis",
  domain = risk_grid$risk_normalized,
  na.color = NA
)

# Create interactive leaflet map
leaflet(risk_grid) %>%
  addTiles(urlTemplate = "https://{s}.tile.openstreetmap.org/{z}/{x}/{y}.png",
           attribution = '© OpenStreetMap contributors') %>%
  addRectangles(
    lng1 = ~lon_min,
    lng2 = ~lon_max,
    lat1 = ~lat_min,
    lat2 = ~lat_max,
    color = ~viridis_pal(risk_normalized),
    fillOpacity = 0.7,
    weight = 0.5,
    popup = paste(
      "Latitude:", ~round(grid_lat, 2), "<br>",
      "Longitude:", ~round(grid_lon, 2), "<br>",
      "Risk Score:", ~round(risk_score, 2), "<br>",
      "Point Count:", ~point_count, "<br>",
      "Avg Confidence:", ~round(avg_confidence, 3)
    ),
    highlightOptions = highlightOptions(
      color = "white",
      weight = 2,
      bringToFront = TRUE
    )
  ) %>%
  addLegend(
    pal = viridis_pal,
    values = ~risk_normalized,
    title = "Tick Risk Score",
    position = "bottomright",
    labels = c("Low", "High")
  ) %>%
  setView(lng = 0, lat = 20, zoom = 2)
```







