---
title: "ml_intro"
author: "Zo Kupelian"
date: "2026-08-31"
output: pdf_document
---

#Option 1

```{r}
#install.packages("tidymodels")
#install.packages("ranger")
library(tidymodels)
library(ranger)
```

split for test and train
```{r}
#for reproducability

set.seed(94)

#initial_split is a useful way to take a chunk for training
data_split <- initial_split(iris, prop = 0.8, strata = Species)

train_data <- training(data_split)
test_data <- testing(data_split)
```

Going to try to predict species based on all of the other variables in iris
```{r}
iris_recipe <- recipe(Species ~ ., data = train_data) %>%
  step_normalize(all_numeric_predictors()) #centering and scaling numeric data
```

```{r}
#defining a random forest classification model

rf_spec <- rand_forest(trees = 100) %>%
  set_engine("ranger") %>%
  set_mode("classification")
```

Creating machine learning workflow
```{r}
iris_workflow <- workflow () %>%
  add_recipe(iris_recipe) %>%
  add_model(rf_spec)
```

train and evaluate model
```{r}
final_fit <- last_fit(iris_workflow, data_split)

collect_metrics(final_fit)
```

```{r}
test_predictions <- collect_predictions(final_fit)
test_predictions %>%
  conf_mat(truth = Species, estimate = .pred_class)
```

#Option 2

following the guide here:

https://shittuolumide.medium.com/machine-learning-basics-building-your-first-predictive-model-in-r-010ffe88d6e4

This uses caret and tidyverse, which implies that it might be updated to use tidymodels instead

Going to use mtcars for the first attempt, then try with iris later

```{r}
#install.packages("tidyverse")
#install.packages("caret")
library(tidyverse)
library(caret)
```

```{r}
data <- mtcars
data

#look for missingness
sum(is.na(data))

#summary stats for all the columns
summary(data)

#split the data into 80% training and 20% testing
train_index <- createDataPartition(data$mpg, p = 0.8, list = FALSE)
train_data <- data[train_index, ]
test_data <- data[-train_index, ]

#train the model, using the train() function from caret
#this is a linear regression model, hence the use of "lm"

model <- train(
  mpg ~ wt + hp, #forumula, mpg as a function of wt and hp
  data = train_data,
  method = "lm",
  trControl = trainControl(method = "cv", number = 5) #five-fold cross validation
)

print(model)
```

Evaluating the model
```{r}
#make predictions
predictions <- predict(model, newdata = test_data)

results <- data.frame(
  Actual = test_data$mpg,
  Predicted = predictions
)

#root mean square error (RMSE)
rmse <- sqrt(mean((results$Actual - results$Predicted)^2))
print(paste("RMSE:", rmse))
```

Not bad! Let's graph it for fun
```{r}
plot(test_data$mpg, predictions, main = "Actual vs. Predicted",
     xlab = "Actual mpg", ylab = "Predicted mpg")
abline(0, 1, col = "red")
```

##recreate
Okay, let's see if I understand this enough to recreate if for iris

```{r}
sum(is.na(iris))
summary(iris)

train_iris_index <- createDataPartition(iris$Petal.Length, p = 0.8, list = FALSE)
train_iris_data <- iris[train_iris_index, ]
test_iris_data <- iris[-train_iris_index, ]
```

```{r}
#training a model to predict petal length given sepal width and length

iris_model <- train(
  Petal.Length ~ Sepal.Width + Sepal.Length,
  data = train_iris_data,
  method = "lm",
  trControl = trainControl(method = "cv", number = 5) #this specifically I need to clarify for myself
)

print(iris_model)
```

```{r}
iris_predictions <- predict(iris_model, newdata = test_iris_data)

iris_results <- data.frame(
  Iris_Actual = test_iris_data$Petal.Length,
  Iris_Predicted = iris_predictions
)

iris_rmse <- sqrt(mean((iris_results$Iris_Actual - iris_results$Iris_Predicted)^2))

print(paste("RMSE:", iris_rmse))
```

Visualize!
```{r}
plot(test_iris_data$Petal.Length, iris_predictions, main = "Actual vs. Predicted Petal Length",
     xlab = "Actual petal length",
     ylab = "Predicted petal length")
abline(0, 1, col = "red")
```

##real data

Okay, looks like I can. That's pretty cool. Maybe I'll give it a shot with the app data.

```{r}
#install.packages("xlsx")
library(xlsx)
library(tidyverse)
library(maps)
library(ggplot2)
library(dplyr)
```

```{r}
app_data <- read.xlsx("app_data.xlsx", sheetIndex = 1)
app_data <- app_data %>%
  select(created_at_utc, latitude, longitude, result, readable, confidence, test_line, invalid_reason, user_action, reasoning, parsed_result, parsed_confidence, parsed_control_line, parsed_test_line, parsed_invalid_reason)

app_data
```

##exploration
```{r}
world_map <- map_data("world")

ticks <- data.frame(lon = as.numeric(c(app_data$longitude)), lat = as.numeric(c(app_data$latitude)))

world_map_graphed <- ggplot() +
  geom_polygon(data = world_map, aes(x = long, y = lat, group = group), fill = "grey", color = "white", linewidth = 0.2) +
  geom_point(data = ticks, aes(x = lon, y = lat), color = "red", size = 1)
world_map_graphed
```

Wow, that's kind of interesting, I wasn't expecting to have some app users in India and Singapore

They'll make for some interesting outliers.

I'll test just using latitude and longitude for now, but I'll set up weather pulls from an API later. Also figure out a way to include date, with later/sooner as the options

##data cleaning
```{r}
#invalids are supposed to be such a small percentage of the dataset in the future that removing them shouldn't impact the model
#though, with this starting dataset they are a kind of huge portion

sum(app_data$result == "invalid", na.rm = TRUE) #76 out of 182 samples are invalid. yikes.

#filtering out invalid results
app_data_clean <- app_data %>%
  filter(!grepl("invalid", result, ignore.case = TRUE)) %>%
  mutate(across(where(is.character), ~ str_replace_all(., c("positive" = "1", "negative" = "0"))))

app_data_clean$result <- as.numeric(app_data_clean$result)
app_data_clean$latitude <- as.numeric(app_data_clean$latitude)
app_data_clean$longitude <- as.numeric(app_data_clean$longitude)
app_data_clean
```


```{r}
set.seed(911)

train_tick_index <- createDataPartition(app_data_clean$result, p = 0.8, list = FALSE)
train_tick_data <- app_data_clean[train_tick_index, ]
test_tick_data <- app_data_clean[-train_tick_index, ]
```

```{r}
tick_model <- train(
  result ~ latitude + longitude,
  data = train_tick_data,
  method = "lm",
  trControl = trainControl(method = "cv", number = 5) #figure out what this is!!
)

print(tick_model)
```

```{r}
tick_predictions <- predict(tick_model, newdata = test_tick_data)

tick_results <- data.frame(
  Tick_Actual = test_tick_data$result,
  Tick_Predicted = tick_predictions
)

tick_rmse <- sqrt(mean((tick_results$Tick_Actual - tick_results$Tick_Predicted)^2))
print(paste("RMSE:", tick_rmse))
```

```{r}
plot(test_tick_data$result, tick_predictions, main = "Actual vs. Predicted Testing Result",
     xlab = "Actual test result",
     ylab = "Predicted test result"
)
abline(0, 1, col = "blue")
```

So! The result of this model isn't great! But I'd have been unbelievably 
surprised if it was considering it's using linear regression for a binary,
spatial set of data.

Next steps are going to be swapping "lm" out for something else, and getting 
weather data to provide that model.

##take 2

```{r}
#set result to a factor for glm with caret, this is a classification model
app_data_clean$result <- as.factor(app_data_clean$result)

set.seed(911)

#80% training, 20% testing. Mostly bc the number of data points we have is small

#stratified by result, maintaining the same class proportion in both sets so we
#don't end up with a testing set full of negatives

train_tick_index <- createDataPartition(app_data_clean$result, p = 0.8, list = FALSE)

#all this is is an index of row numbers for the training set. Just numbers.

train_tick_data <- app_data_clean[train_tick_index, ]
test_tick_data <- app_data_clean[-train_tick_index, ]
```

```{r}
#uses caret::train() function to train the model, may want to look at alternatives

#result (positive or negative) predicted by latitude and longitude, using only
#geographic coordinates to classify the result

tick_model <- train(
  result ~ latitude + longitude,
  data = train_tick_data,
  method = "glm", #generalized linear regression model
  family = binomial(link = "logit"), #specifying logistic regression, binary outcome and outputting probs 0 to 1
  trControl = trainControl(method = "cv", number = 5) #5-fold cross-validation during training
)

#What does that trControl mean? It divides the training data into 5 subsets, trains 5 times, each time using 4 folds and 1 to validate
#you can increase this, but the more subsets you divide the training data into, the smaller they get. Diminishing returns and all that.

print(tick_model)
#coefficients, performance metrics, sample sizes, etc.
```

An accuracy of 62.8% percent with the 5-fold validation isn't amazing, but it's 
better than flipping a coin!

Below is predictions and confusion matrix
```{r}
#these are probability predictions on the test set, taking the test data and 
#using it to predict "result = 1" or "result = 0"

#eventually we'll compare and see if it matches up to the *actual* "result = 1" or "result = 0"

#P(result = 1 | latitude, longitude)

tick_predictions_prob <- predict(tick_model, newdata = test_tick_data, type = "prob") #type = "prob" is probabilities for each class
tick_predictions <- predict(tick_model, newdata = test_tick_data, type = "raw") #0 or 1 is returned as the predicted class

#making that dataframe of the actual test data and the predicted test data
tick_results <- data.frame(
  Tick_Actual = test_tick_data$result,
  Tick_Predicted = tick_predictions
)

#confusion matrix to show true positives, true negatives, false positives, false 
#negatives, accuracy, sensitivity, specificity, etc.
confusionMatrix(tick_predictions, test_tick_data$result)
```

70% accuracy, so not terrible. Good sensitivity, specificity could be better.

```{r}
#this is visualization

# the `1` extracts the probability column for the positive class

tick_results <- data.frame(
  Tick_Actual = test_tick_data$result,
  Tick_Predicted = tick_predictions,
  Probability = tick_predictions_prob$`1`  # Predicted probability
)

# Confusion Matrix Visualization

#x axis shows the actual results, the fill colors is the predicted results

ggplot(tick_results, aes(x = Tick_Actual, fill = Tick_Predicted)) +
  geom_bar(position = "dodge") +
  labs(title = "Actual vs. Predicted Classifications",
       x = "Actual Result", y = "Count") +
  theme_minimal()
```

The model does a great job predicting when ticks are going to be negative based
on location data. It does a less great job when ticks are going to be positive
based on location data.
