---
title: "US_cases"
author: "Zo Kupelian"
date: "2026-08-06"
output: pdf_document
---

#libraries
```{r}
#install.packages("tidycensus")
library(ggplot2)
library(dplyr)
library(tidyverse)
library(tidyr)
library(tigris)
library(sf)
library(stringr)
library(magick)
library(tidycensus)
library(httr)
library(readxl)
library(readr)
```

#load'n'look
```{r}
#this was available on the DHHS website, but it'll be uploaded to the repo
US_cases <- read.csv("US_case_incidence.csv")

#need to manually remove all of the commas from the data cells
US_cases$X2008 <- gsub(",", "", US_cases$X2008)
US_cases$X2009 <- gsub(",", "", US_cases$X2009)
US_cases$X2010 <- gsub(",", "", US_cases$X2010)
US_cases$X2011 <- gsub(",", "", US_cases$X2011)
US_cases$X2012 <- gsub(",", "", US_cases$X2012)
US_cases$X2013 <- gsub(",", "", US_cases$X2013)
US_cases$X2014 <- gsub(",", "", US_cases$X2014)
US_cases$X2015 <- gsub(",", "", US_cases$X2015)
US_cases$X2016 <- gsub(",", "", US_cases$X2016)
US_cases$X2017 <- gsub(",", "", US_cases$X2017)
US_cases$X2018 <- gsub(",", "", US_cases$X2018)
US_cases$X2019 <- gsub(",", "", US_cases$X2019)
US_cases$X2020 <- gsub(",", "", US_cases$X2020)
US_cases$X2021 <- gsub(",", "", US_cases$X2021)
US_cases$X2022 <- gsub(",", "", US_cases$X2022)
US_cases$X2023 <- gsub(",", "", US_cases$X2023)

#make it numeric
US_cases <- US_cases %>%
  mutate(
    X2008 = as.numeric(X2008),
    X2009 = as.numeric(X2009),
    X2010 = as.numeric(X2010),
    X2011 = as.numeric(X2011),
    X2012 = as.numeric(X2012),
    X2013 = as.numeric(X2013),
    X2014 = as.numeric(X2014),
    X2015 = as.numeric(X2015),
    X2016 = as.numeric(X2016),
    X2017 = as.numeric(X2017),
    X2018 = as.numeric(X2018),
    X2019 = as.numeric(X2019),
    X2020 = as.numeric(X2020),
    X2021 = as.numeric(X2021),
    X2022 = as.numeric(X2022),
    X2023 = as.numeric(X2023)
  )

US_cases_longer <- pivot_longer(
    data = US_cases,
    cols = c("X2008", "X2009", "X2010", "X2011", "X2012", "X2013", "X2014", "X2015", "X2016", "X2017", "X2018", "X2019", "X2020", "X2021", "X2022", "X2023"),
    names_to = "Year",
    values_to = "Cases"
  )
head(US_cases_longer)
```


#dataprep for heatmapping
```{r}
#make a folder for these heatmaps, going to reference for the gif later
dir_out <- file.path(".", "US_heatmaps_folder")
dir.create(dir_out, recursive = TRUE)

#just map data
US_map <- states(cb = TRUE, year = 2024) #2024 state borders
US_map <- US_map %>%
  rename("State" = "NAME") %>%
  filter(!State %in% c("United States Virgin Islands", 
                       "Commonwealth of the Northern Mariana Islands", "Guam",   #have to exclude these, no data
                       "American Samoa", "Puerto Rico", "Hawaii")) %>%
  shift_geometry() #makes the map fit better on the plot

#take the Xs out of the Year column, they were useful for renaming until this point
US_cases_longer$Year <- gsub("X", "", US_cases_longer$Year)
US_cases_longer$State <- gsub("\x86", "", US_cases_longer$State, fixed = TRUE, useBytes = TRUE)

#no idea why \x86 is attached to Wisconsin, Pennsylvania, West Virginia, Virginia, Vermont, Hawaii, but removed it now

US_cases_longer
```

#heatmapping function
```{r}
US_year_heatmap <- function(year_choice) {

#year by year, so filter the data by the year the user wants to see
  US_cases_year <- US_cases_longer %>%
    filter(Year == year_choice)

#join
merged_data <- left_join(US_map, US_cases_year, by = "State")
as.data.frame(merged_data)

#standardizing across the entire range of case counts
case_limits <- c(0, 25000) #max cases per year in one state is ~22000, I think

#-------------------------------------------------------------------------------
#remove case limits from above and below if we want to see states compare to
#each other in a certain year

#keep case limits if we're looking across the whole period of time, incidence
#-------------------------------------------------------------------------------

#time to plot
heatmap <- ggplot(merged_data) +
  geom_sf(aes(fill = Cases), color = "white", linewidth = 0.2) +
  scale_fill_viridis_c(limits = case_limits, option = "plasma", name = "Number of Cases") +
  theme_void() +
  labs(title = paste("US Cases per State Heatmap in", year_choice))
print(heatmap)

fp <- file.path(dir_out, paste0("US_case_heatmap_", year_choice, ".png"))

#time to save
ggsave(plot = heatmap, filename = fp, device = "png")

}

lapply(2008:2023, US_year_heatmap)
```

Now, this is just case numbers. I wonder if I can make prevalence maps...

To do that, I'd need state populations for every year between 2008 and 2023, 
which should be doable.

Use # of cases/population per year, and I've got prevalence.

Can't be that hard!

#load in pop data
```{r}
census_api_key("9c6d69c7cf5285a77a6203f0ef22788ba10bd69", install = TRUE, overwrite = TRUE)

#2008-2009 data, since the census stored population data differently across years, for some reason

tmp_xls <- tempfile(fileext = ".xls")
GET("https://www2.census.gov/programs-surveys/popest/tables/2000-2010/intercensal/state/st-est00int-01.xls", write_disk(tmp_xls))

pop_2000s <- read_excel(tmp_xls, skip = 3) %>%
  rename("State" = "...1", "2008_pop" = "2008", "2009_pop" = "2009") %>%
  select("State", "2008_pop", "2009_pop")
pop_2000s
```

```{r}
#2010-2019 data

pop_2010s <- read_csv(
  "https://www2.census.gov/programs-surveys/popest/datasets/2010-2019/state/detail/sc-est2019-alldata6.csv",
  show_col_types = FALSE
)

#this massive data set is super informative. 0 is the combined, 1+ is the different
#categories. For now, I'm just interested in the total number, so I'm only including
#the rows that have 0s in them.

pop_2010s <- pop_2010s %>%
  filter(SEX == 0, ORIGIN == 0) %>% #all sexees and all origins
  group_by(NAME) %>%
  summarise(across(starts_with("POPESTIMATE"), sum, .names = "{.col}")) %>%
  tidyr::pivot_longer(-NAME, names_to = "year_var", values_to = "population") %>%
  mutate(year = readr::parse_number(year_var)) %>%
  select(NAME, year, population) %>%
  arrange(NAME, year)
pop_2010s
```

```{r}
#2020-2023 data

#this one actually summarizes on the state level for SUMLEV == "040", which is
#very nice and convenient instead of processing like the 2010s.

pop_2020s <- read_csv(
  "https://www2.census.gov/programs-surveys/popest/datasets/2020-2023/state/totals/NST-EST2023-ALLDATA.csv",
  show_col_types = FALSE
) %>%
  filter(SUMLEV == "040") %>%
  select(NAME, starts_with("POPESTIMATE")) %>%
  pivot_longer(-NAME, names_to = "year", values_to = "population") %>%
  mutate(year = as.integer(gsub("POPESTIMATE", "", year)))
pop_2020s
```

#wrangle

```{r}
#next step is to wrangle this data into a format that can be combined

pop_2000s$State <- gsub(".", "", pop_2000s$State, fixed = TRUE)
pop_2000s <- pop_2000s %>%
  filter(!State %in% c("United States", "Northeast", "Midwest", "South", "West", 
                       "Hawaii", "Puerto Rico")) %>%
  filter(row_number() <= n()-9) %>%
  rename("2008" = "2008_pop", "2009" = "2009_pop") %>%
  pivot_longer(!State, names_to = "year", values_to = "population")

pop_2010s <- pop_2010s %>%
  rename("State" = "NAME") %>%
  filter(!State %in% c("United States", "Northeast", "Midwest", "South", "West", 
                       "Hawaii", "Puerto Rico"))
pop_2020s <- pop_2020s %>%
  rename("State" = "NAME") %>%
  filter(!State %in% c("United States", "Northeast", "Midwest", "South", "West", 
                       "Hawaii", "Puerto Rico"))

pop_2000s
pop_2010s
pop_2020s
```

Very lovely, now I need to join them in a way that has each state's year.

```{r}
pop_early = rbind(pop_2000s, pop_2010s)
pop_early <- pop_early[order(pop_early$State),]
pop_early

pop = rbind(pop_early, pop_2020s)
pop <- pop[order(pop$State), ]
pop
```

Nice, now pop and US_cases_longer should match up
```{r}
US_cases_state <- US_cases_longer %>%
  filter(!State %in% c("U.S. Total"))

pop
US_cases_state
```







